---
title: "Methods to estimate fishing effort"
author: "ICES WKSSFGEO2 workshop"
date: "2023-03-14"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r functions , eval=FALSE}

#Robin Lovelace code:
lonlat2UTM <- function( lonlat ){

  utm <- (floor((lonlat[1] + 180) / 6) %% 60) + 1

  if( lonlat[2] > 0){

    utm <- utm + 32600

  }else{

    utm <- utm + 32700

  }

  return(utm)

}


#for those wanting to try iapesca package
#iapesca package

https://gitlab.ifremer.fr/iapesca

```

### First, read in and process the data for analysis
Use the `moveHMM` package to compute the distances and turning angles between consecutive location records. Remove records with missing step length, and convert the step length values from meters to knots. 


```{r prep data, , eval=FALSE}

library(moveHMM)
library(sf)
library(vctrs)

df <- read.csv("example_PT_IPMA_MRufino.csv", header=TRUE)

lonlat2UTM(apply(df[,c("Longitude","Latitude")],2,function(x) mean(x,na.rm=TRUE)))

df_sf = st_as_sf(df, coords = c("Longitude", "Latitude"), crs = 4326)
df_sf_utm<-st_transform(df_sf, 32629)
coords_utm<-as.data.frame(st_coordinates(df_sf_utm))#extract coordinates
df$x<-coords_utm$X
df$y<-coords_utm$Y

prepdat <- moveHMM::prepData(df, type = "UTM", coordNames = c("x", "y")) 

prepdat <- prepdat[!is.na(prepdat$step),]
prepdat$speed <- prepdat$step/60*1.9438
hist(prepdat$speed)
```

### 1. "Overall" speed threshold

We use the speed distribution figure generated (histogram) to define reasonable means and standard deviations for each distribution of speed for each of the three underlying behaviours in this fishery. 

We then run an Expectation Maximisation (EM) algorithm to identify the parameters of each of the underlying univariate distribution (hauling, deploying, and steaming) via maximum likelihood estimation.The lowest mean and its associated standard deviation corresponds to hauling. To determine the upper threshold of speeds attained during hauling activities, we use the mean + 1.96*standard deviation. 


```{r overall speed threshold, eval=FALSE}
library(mixtools)

EM_all <- normalmixEM(prepdat$speed, mu = c(1,6,10), sigma =c( 1,1,1))#use histogram of speeds of all trips-eyeball starting values
plot(EM_all, density=TRUE,loglik=FALSE)#check distributions
 
prepdat$speed_threshold_EM <- EM_all$mu[1]+1.96*EM_all$sigma[1]
#assign a maximum value for the speed value for hauling - mean + 1.96 SD from the distribution identified by the EM algorithm

prepdat$overall_EM_beh <- ifelse(prepdat$speed < prepdat$speed_threshold_EM, "hauling", "not_hauling")
```

### 2. Trip-based Expectation Maximization

Here we split the data into individual fishing trips and then use the EM algorithm to estimate means and standard deviations for each of the three behaviours (hauling, deopoyment, and steaming) for each individual trip.To determine the upper threshold of speeds attained during hauling activities, we use the mean + 1.96*standard deviation for each trip. 

```{r trip_based EM, eval=FALSE}
library(dplyr)

lst <- split(prepdat, prepdat$ID) #divide per trip - 8 trips
List <- list()
N <- length(lst)

for (i in 1:N){

trip_EM <- normalmixEM(lst[[i]]$speed, mu = c(1,4,8), sigma =c( 1,1,1))
ks <- (unlist(c(trip_EM$lambda,trip_EM$mu, trip_EM$sigma,trip_EM$loglik)))
ks <- as.data.frame(t(ks))
List[[length(List)+1]] <- ks

if (sum(ks) != 0 ) {

  print(paste("i=",i))
  plot(trip_EM, density=TRUE,loglik=FALSE)

  } else {
      
      (paste("i=",unique(trip_EM[[i]]$ID), "is 0. Not processed")) # Print out progress
      next 
  } 
}

trip_EM_par <- bind_rows(List)
colnames(trip_EM_par) <- c("lambda1", "lambda2", "lambda3","mu1", "mu2", "mu3","sd1", "sd2", "sd3","loglik")

# add trip ID to trip_EM parameter dataframe
f <- NULL
N <- length(lst)
for (i in 1:N){
  f <- rbind(f,data.frame(ID = unique(lst[[i]]$ID)))
}
trip_EM_par$ID <- f$ID

prepdat <- merge(prepdat, trip_EM_par[,c("ID","mu1","sd1")], by=c("ID"))
prepdat$speed_threshold_trip_EM <- prepdat$mu1 + 1.96 * prepdat$sd1
prepdat$trip_EM_beh <- ifelse(prepdat$speed < prepdat$speed_threshold_trip_EM, "hauling", "not_hauling")

names(prepdat) #remove unnecesary columns
prepdat <- prepdat %>% 
  select(-c("speed_threshold_EM","mu1","sd1","speed_threshold_trip_EM"))
```



### 3. Binary Clustering for behavioural annotation using Gaussian mixture models on a trip-by-trip basis
The third approach we use is also a Gaussian mixture model but 

```{r trip based EMbc, eval=FALSE}
# load EMbC package
library(EMbC)

prepdat$date <- as.POSIXct((prepdat$date), format = "%Y-%m-%d %H:%M:%S")

# transform UTM coordinates to WGS84 (lon, lat)
xy <- data.frame(X = prepdat$x, Y = prepdat$y)
library(sp)
 coordinates(xy) <- c("X", "Y")
 proj4string(xy) <- CRS('+proj=utm +zone=30 +datum=WGS84 +units=m 
                        +no_defs +ellps=WGS84 +towgs84=0,0,0')  # for example
 coor <- as.data.frame(spTransform(xy, CRS("+proj=longlat +datum=WGS84")))
prepdat$lon <- coor$X
prepdat$lat <- coor$Y

# choose relevant columns for analysis: time stamp, x and y
names(prepdat)
prepdat3 <- prepdat %>%
  select("ID","date","lon","lat")

EMbc_value <- data.frame() 
out.list3 <- split(prepdat3, prepdat3$ID)

# loop through the trips in out.list2, by name
for (ii in names(out.list3)){
    iidf <- as.data.frame(out.list3[[ii]][,2:4])
    # carry out speed/turn bivariate binary clustering for each trip, using stbc()
    EMbc <- stbc(iidf, info=-1)
    # take the numeric vector of clustering labels (A) and turn them into a data frame
    EMbcA <- as.data.frame(EMbc@A)
    # attach the trip ID to the data frame
    EMbcA$trip_ID <- ii
# plot clustering visually
sctr(EMbc)
# plot track with labels
view(EMbc)
# print trip label for sense-check
print(ii)
# attach trip-specific output to main outout data frame 
EMbc_value <- rbind(EMbc_value, EMbcA)
} 

dev.off()
graphics.off()

# hauling events are associated with low speeds, low turning angles
EMbc_value$Embc_simple <- ifelse(EMbc_value$`EMbc@A`==1 | EMbc_value$`EMbc@A`==2, 
                                 "hauling", "not_hauling")

# attach results to the main data set and check it looks ok
prepdat$trip_EMbc_beh <- EMbc_value$Embc_simple
head(prepdat)
```


### 4. Hidden Markov model with speed only 
The next two approaches employ hidden Markov models implemented in the `R` package `moveHMM`. 
We first fit a model to speed only. We use parameters from trip-based Gaussian mixture model fitted in 2. above, as initial parameters for gamma distributions of step length (distance travelled between regular observations in time).


```{r HMM speed, eval=FALSE}
# load the moveHMM and dplyr package
library(moveHMM)
library(dplyr)

# split the data into trip (total 5 trips)
in.list4 <- split(prepdat, prepdat$ID)  
out.list4 <- list()
N <- length(in.list4)

# loop through trips to find starting values using a Gaussian mixture model 
# (as in 2, above) to do so
for (i in 1:N){

trip_EM <- normalmixEM(in.list4[[i]]$step, mu = c(20,150,300), sigma =c(20,20,20))
ks <-(unlist(c(trip_EM$lambda, trip_EM$mu, trip_EM$sigma, trip_EM$loglik)))
ks <- as.data.frame(t(ks))
out.list4[[i]] <- ks

if (sum(ks) != 0 ) {

  print(paste("i=",i))

  } else {
      
      (paste("i=",unique(trip_EM[[i]]$ID), "is 0. Not processed")) # Print out progress
      next 
  } 
}

# combine the out.list4 object back into a dataframe
trip_EM_par_step <- bind_rows(out.list4)
head(trip_EM_par_step)
colnames(trip_EM_par_step) <- c("lambda1", "lambda2", "lambda3",
                                "mu1", "mu2", "mu3", 
                                "sd1", "sd2", "sd3",
                                "loglik")

# add trip ID to trip_EM_par_step dataframe
trip_EM_par_step$ID <- unique(prepdat$ID)

# now fit HMM using the trip-specific GMM estimates for speed as starting values

library(purrr)

# I have to create a moveData object again - but duplicate remove step and angle
in.list4hmm <- split(prepdat, prepdat$ID)
out.list4hmm <- map(in.list4hmm, function(x) select(x, -c(step, angle, id))) %>% 
  map(., function(x) prepData(x, type = "UTM", coordNames = c("x", "y"))) 
head(out.list4hmm[[1]])

N <- length(out.list4hmm)

for (i in 1:N){
# to initiate numerical maximization (optimization of the likelihood) starting values 
# for the model parameters need to be specified.

  mu0 <- c(trip_EM_par_step$mu1[i], 
              trip_EM_par_step$mu2[i], 
              trip_EM_par_step$mu3[i])
  sigma0 <- c(trip_EM_par_step$sd1[i],  
             (trip_EM_par_step$sd2[i]-trip_EM_par_step$sd2[i]/4),
             (trip_EM_par_step$sd3[i]+2)) 
  
  # we specify a smaller SD for speed during deployment behaviour than during steaming 

stepPar0 <- c(mu0,sigma0)
stepDist <- "gamma"

# fit an HMM with 3 states
m <- fitHMM(data=out.list4hmm[[i]], nbStates=3, stepPar0=stepPar0, 
            verbose=0, stepDist=stepDist, angleDist='none') 
# print(m)
ks <- as.data.frame(unlist(m$mle))
vit <- viterbi(m)

if (sum(ks) != 0 ) {
  write.table(ks, paste0(trip_EM_par_step$ID[i],".txt"))
  write.table(vit, paste0('viterbi_speed',trip_EM_par_step$ID[i],".txt"))
  print(paste("i=",i))
  plot(m,compact=TRUE,ask=FALSE)#the compact bit puts all vessels in one graph

# compute the pseudo-residuals
#pr <- pseudoRes(m)
# time series, qq-plots, and ACF of the pseudo-residuals
#plotPR(m)

  } else {
      
      print(paste("i=",i, "is 0. Not processed")) # Print out progress
      next 
    } 
}

dev.off()
graphics.off()

list.filenames <- list.files(path = ".", full.names = FALSE, pattern="^viterbi_speed")
datalist <- map(list.filenames, function(x) read.table(x, header=T, sep=""))
names(datalist) <- list.filenames
trial <- do.call(rbind.data.frame, datalist)
prepdat$viterbi_speed <- trial$x #2221 obs

prepdat$trip_HMM_speed_beh <- ifelse(prepdat$viterbi_speed==1, "hauling", "not_hauling")
```

### 5. Hidden Markov Model with speed and turning angle
In the second HMM approach we fit a model to speed and turning angle. We use parameters from trip-based Gaussian mixture model fitted in 2. above, as initial parameters for the mean of the gamma distributions of step length (distance travelled between regular observations in time). We provided a smaller standard deviation for hauling and deploying gear than for steaming. For the turning angle we chose starting values to reflect that we expected hauling to have a larger turning angle than steaming.

```{r HMM speed and angle, eval=FALSE}
# load the moveHMM package
library(moveHMM)

# split the data into trip (total 5 trips)
in.list5hmm <- split(prepdat, prepdat$ID)  
out.list5hmm <- map(in.list5hmm, function(x) select(x, -c(step, angle, id))) %>% 
  map(., function(x) prepData(x, type = "UTM", coordNames = c("x", "y"))) 
N <- length(in.list5hmm)

for (i in 1:N){

  mu0 <- c(trip_EM_par_step$mu1[i], 
           trip_EM_par_step$mu2[i], 
           trip_EM_par_step$mu3[i])
  sigma0 <- c(trip_EM_par_step$sd1[i], 
             (trip_EM_par_step$sd2[i]-trip_EM_par_step$sd2[i]/4), 
             (trip_EM_par_step$sd3[i]+2))

  stepPar0 <- c(mu0,sigma0)
  angleMean0 <- c(pi,pi/2,0) # angle mean 
  kappa0 <- c(0.1,0.2,0.8) # angle concentration
  anglePar0 <- c(angleMean0,kappa0)
  stepDist <- "gamma"
  angleDist <- "wrpcauchy"

  m <- fitHMM(data=out.list5hmm[[i]], nbStates=3, stepPar0=stepPar0, 
              verbose=0, stepDist=stepDist, anglePar0=anglePar0, angleDist=angleDist) 
  ks <- as.data.frame(unlist(m$mle))
  vit <- viterbi(m)

if (sum(ks) != 0 ) {
  write.table(ks, paste0(trip_EM_par_step$ID[i],".txt"))
  write.table(vit, paste0('viterbi_angle',trip_EM_par_step$ID[i],".txt"))
  print(paste("i=",i))
  plot(m, compact=TRUE, ask=FALSE) # the compact=TRUE puts all vessels in one graph

  } else {
      
      print(paste("i=",i, "is 0. Not processed")) # print out progress
      next 
    } 
}

dev.off()
graphics.off()
 
list.filenames <- list.files(path = ".", full.names = FALSE, pattern="^viterbi_angle")

datalist <- map(list.filenames, function(x) read.table(x, header=T, sep=""))
names(datalist) <- list.filenames
trial <- do.call(rbind.data.frame, datalist)
prepdat$viterbi_speed_angle <- trial$x
prepdat$trip_HMM_speed_angle_beh <- ifelse(prepdat$viterbi_speed_angle==1, 
                                           "hauling", "not_hauling")

names(prepdat)
prepdat <- prepdat %>% select(-c("viterbi_speed","viterbi_speed_angle"))

# check number of correctly assigned behaviours for each position

names(prepdat)
prepdat <- prepdat %>% 
  mutate(behaviour_simple = ifelse(prepdat$behaviour=="hauling", 
                                   "hauling", "not_hauling")) 

prepdat <- prepdat %>%
    mutate(match_overall_EM_beh = ifelse(
            prepdat$overall_EM_beh==prepdat$behaviour_simple, 1,0),
           match_trip_EM_beh = ifelse(
            prepdat$trip_EM_beh==prepdat$behaviour_simple,  1,0),
           match_trip_EMbc_beh = ifelse(
            prepdat$trip_EMbc_beh==prepdat$behaviour_simple, 1,0),
           match_trip_HMM_speed_beh = ifelse(
            prepdat$trip_HMM_speed_beh==prepdat$behaviour_simple, 1,0),
           match_trip_HMM_speed_angle_beh = ifelse(
            prepdat$trip_HMM_speed_angle_beh==prepdat$behaviour_simple, 1,0))

sum(prepdat$match_overall_EM_beh) #2000
sum(prepdat$match_trip_EM_beh) #2021
sum(prepdat$match_trip_EMbc_beh) #2039
sum(prepdat$match_trip_HMM_speed_beh) #2004
sum(prepdat$match_trip_HMM_speed_angle_beh) #2028
```

